Supplemental Information: Aligning marine species range data to better serve science and conservation

### 2016-09-09 ms_3_res_disc_concl.Rmd

@@

## Distribution and range area alignment

[discussion] To explore differences in species distribution and range between the two datasets, we assessed the distribution alignment (how much of the smaller range falls within the larger range, i.e., where on the map) relative to the area ratio (ratio of smaller range area to larger range area, i.e., how much of the map) for each shared species (Fig 2A).  Where two independent methods predict presence (or absence) of a species in the same location, we can assume a higher confidence in the accuracy of that prediction.  Where one method predicts presence while the other predicts absence, we can investigate potential sources of commission and omission errors to refine our understanding of the species' range.  Understanding these sources of error can provide guidance to improve predictions of species range even when only one source of spatial data is available.

@@

[discussion] This analysis revealed a weak negative linear pattern, suggesting that increasing similarity in range area correlates with decreasing distribution alignment.
 AquaMaps tends to extrapolate species ranges into suitable areas beyond known occurrences, such that each additional unit of range predicted by AquaMaps will fall in different locations than an additional unit of range predicted using IUCN methodology. For species with dissimilar range areas, predicted distribution for the smaller range can more easily fall within the generous bounds of the larger range. For species with increasingly similar range areas, differences in methodology become more difficult to "hide," and the distribution alignment generally becomes poorer.

[results] The mean distribution alignment for species included in both datasets was `r dist_align_mean`%; the mean area alignment was `r area_align_mean`%. By dividing the paired map species into quadrants based on these means, we highlight categories of relationships that help further explain this general pattern. Representative maps from each category are provided in the supporting materials (Fig S1).

@@

[discussion] Noting from Fig 2B that corals dominate the upper-left "distribution-aligned" quadrant of Fig 2A (n = `r n_q2_corals`; `r round(n_q2_corals/n_q2 * 100, 1)`% of all species in this quadrant), we chose to examine the effect of explicitly restricting IUCN ranges to depths based on species' life histories.  Corals offer an excellent case study, due to their foundational role in biodiverse habitats, as well as their lack of mobility and reliance on photosynthesis.  Ocean depth preference is explicitly included in AquaMaps models.  While depth is recommended by the IUCN as a criterion for providers of range maps ("The limits of distribution can be determined by using known occurrences of the species, along with the knowledge of habitat preferences, remaining suitable habitat, elevation limited, and other expert knowledge of the species and its range." [@REF http://www.iucnredlist.org/technical-documents/red-list-training/iucnspatialresources]), it is not presented as a requirement, so we cannot take its inclusion for granted.

>>>>

[brief overview of results/discussion?]

@@

## Distribution and range size alignment

[discussion] [results] Comparing distribution alignment and area ratio for the `r format(n_spp_both, big.mark = ',')` paired map species (Fig 2A), a weak negative linear pattern appears to emerge, suggesting that increasing similarity in range area correlates very slightly with decreasing distribution alignment (R^2 = .016).  The pattern itself is unimportant, and emerges simply due to the nature of the analysis and the datasets:  AquaMaps tends to extrapolate species ranges into suitable areas beyond known occurrences more frequently than IUCN maps, such that each additional unit of range predicted by AquaMaps will fall in different locations than an additional unit of range predicted using IUCN methodology. For species with dissimilar range areas, predicted distribution for the smaller range can more easily fall within the generous bounds of the larger range. For species with increasingly similar range areas, differences in methodology become more difficult to "hide," and the distribution alignment generally becomes poorer.

The mean distribution alignment for species included in both datasets was `r dist_align_mean`%; the mean area alignment was `r area_align_mean`%. By dividing the paired map species into quadrants based on these means, we highlight categories of relationships that help further explain this general pattern. Representative maps from each category are provided in the supporting materials (Fig S1).

Where two independent methods predict presence (or absence) of a species in the same location, we can assume a higher confidence in the accuracy of that prediction.  Where one method predicts presence while the other predicts absence, we can investigate potential sources of commission and omission errors to refine our understanding of the species' range.  Understanding and addressing these sources of error provides guidance to improve predictions of species range even when only one source of spatial data is available.

@@

@@

[discussion] Noting from Fig 2B that corals dominate the upper-left "distribution-aligned" quadrant of Fig 2A (n = `r n_q2_corals`; `r round(n_q2_corals/n_q2 * 100, 1)`% of all species in this quadrant), we chose to examine the effect of explicitly restricting IUCN ranges to depths based on species' life histories.  Corals offer an excellent case study, due to their foundational role in biodiverse habitats, as well as their lack of mobility and reliance on photosynthesis.  Ocean depth preference is explicitly included as a parameter in AquaMaps models.  While depth is recommended by the IUCN as a criterion for providers of range maps ("The limits of distribution can be determined by using known occurrences of the species, along with the knowledge of habitat preferences, remaining suitable habitat, elevation limited, and other expert knowledge of the species and its range." [@REF http://www.iucnredlist.org/technical-documents/red-list-training/iucnspatialresources]), it is not presented as a requirement, so we cannot take its inclusion for granted.

### 2016-09-09 ms_5_figures.Rmd

___NOTE:___ These images are all based on .png outputs, mostly from .psd files; submit the .tif versions for publication.

@@

![](figures/Fig_1.png)

@@

![](figures/Fig_2.png)

>>>>

___NOTE:___ These images are all based on .png outputs, exported from .psd files; submit the .tif versions for publication.

@@

![Taxonomic and geographic representation](figures/Fig_1.png)

@@

![Quadrant plot and spp breakdown by quadrant](figures/Fig_2.png)

### 2016-09-09 ms_6_supplemental.Rmd

methods:
IUCN raw -> grid -> raster -> final
AM raw -> presence
bathymetry poly -> raster?

@@

__Fig S3.__ Representative species maps to illustrate each quadrant from Fig 2A.  Each map is positioned to match its quadrant in Fig 2A.
(A) Distribution-matched: Conus episcopatus, the dignified cone snail.  Ranges shows excellent overlap in the western Pacific, though IUCN range extends beyond the bounds of the AquaMaps range.
(B) Well-matched: Kajikia albida, the Atlantic white marlin.  Ranges from each data set show nearly complete overlap, and very similar area of coverage.
(C) Poorly-matched: Acanthurus nigroris, the blue-lined surgeonfish.  IUCN predicts species range only near the Hawaiian islands; AquaMaps predicts extensive range throughout the central and western Pacific Ocean.  The datasets align in neither distribution nor area of coverage.
(D) Area-matched: Conus magnificus, the magnificent cone snail.  Ranges overlap in the southern Pacific, but align poorly elsewhere.  The total area for each range is similar.

>>>>

methods:
IUCN raw -> grid -> raster -> final
AM raw -> presence

@@

__Fig S3.__ Representative species maps to illustrate each quadrant from Fig 2A.  Each map is positioned to match its quadrant in Fig 2A.
(A) Distribution-matched: Conus episcopatus, the dignified cone snail.  Distributions shows excellent overlap in the western Pacific, though IUCN range extends well beyond the bounds of the AquaMaps range.
(B) Well-matched: Kajikia albida, the Atlantic white marlin.  Distributions from each data set show nearly complete overlap, and very similar range size.
(C) Poorly-matched: Acanthurus nigroris, the blue-lined surgeonfish.  IUCN predicts species distribution only near the Hawaiian islands; AquaMaps predicts extensive distribution throughout the central and western Pacific Ocean.  The datasets align in neither distribution nor range size.
(D) Area-matched: Conus magnificus, the magnificent cone snail.  Distributions overlap in the southern Pacific, but align poorly elsewhere.  The range sizes are similar.
